Asymmetric relationships and their effects on coexistence

### 2023-07-19 paper/Paper.Rnw
started a general effort to trim and make more accessible, but decided I might want to wait until after discussing Jasmin's recent explanatory efforts before proceding

%Currently 150 of an allowed 150 words for Ecology Letters
\begin{abstract}
\noindent Species coexistence attracts wide 
interest in ecology. Modern coexistence
theory (MCT) identifies coexistence mechanisms,
one of which, storage effects, hinges on relationships between fluctuations
in environmental and competitive pressures. However, such relationships
are typically measured using covariance, which does not account for 
the possibility that environment and competition may be more related to each other
when they are strong than when weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATAs) are common
between ecological variables, and are important for 
extinction risk, ecosystem stability, and other phenomena. 
We extend the MCT, decomposing storage effects to show the influence of ATAs.
Analysis of a simple model
and an empirical example using diatoms
illustrate that ATA influences can be comparable in magnitude to other
mechanisms of coexistence, and that ATAs can make the 
difference between species coexistence and competitive exclusion. 
ATA influences may be an important new mechanism of coexistence. 
\end{abstract}

@@

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships and to understand how two variables interact, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has previously been referred to as 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are extremely common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argue that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting the potential for rescue effects; 
and \cite{Ghosh_2021} show that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; and many other examples have also been previously explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition and coexistence have not yet been investigated. Doing so is the overarching goal of this study.

@@

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence species coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition, coexistence, and the maintenance of biodiversity, by means of the 
following specific goals. 
G1) We will augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
We examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) We will explore, using the simple lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and to determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence and biodiversity maintenance
are likely to be negligible or meaningfully large. 
G3) We will provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs relative to other mechanisms of coexistence.
The example is a diatom system in a chemostat. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition and coexistence.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance.

@@

To evaluate the importance of ATAs for a given scenario,
one can quantify
the magnitude of the new ATA effect term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms for the 
scenario.  One typically (but see below) must examine this quantity for each species, $i$,
to obtain complete information about coexistence, since coexistence requires both species populations to 
be able to grow when rare. We also defined scenarios of ``ATA rescue''
and ``ATA exclusion''. \emph{ATA rescue} was considered to have occurred for a given scenario if coexistence
was possible when ATAs were taken into account, i.e., GWR$>0$ for both species, but impossible otherwise, i.e.,
the sum of the non-ATA terms in (\ref{eq:full_decomp}) was negative for at least one species. \emph{ATA exclusion} was 
considered to have occurred if the species did not coexist when ATAs
were taken into account, i.e., GWR$<0$ for at least one species, but would have coexisted were it not for ATAs, 
i.e., the sum of the non-ATA terms in (\ref{eq:full_decomp})
was positive for both species.

@@

%Setup - two versions, and basics which apply to both
To achieve goal G2 of the Introduction, we applied the framework of Theory to the lottery model, 
comparing the magnitude of the ATA effect on coexistance, $\Delta_i^{[EC]}$, to the magnitudes 
of the other terms of 
the decomposition (\ref{eq:full_decomp}) for various model parameterizations. We used two versions 
of the lottery model which we now specify. The versions both used the same dynamical equations
(Introduction) and assumed temporally independent and identically distributed (iid) fecundities, but the
versions differed in the distributions of $(B_1,B_2)$ used. Both distributions were based on 
transformations of the distributions $(b_1,b_2)$ pictured in Fig. \ref{fig:pedagog1}a-c;
the details of how the left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c), 
and symmetrically (Fig. \ref{fig:pedagog1}b) tail associated alternatives for $(b_1,b_2)$ were generated 
are in SI section \ref{SI-sect:noise}.

@@

%Differences with the original, analytic version of the MCT
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to expand the mathematics of the original MCT to consider ATAs.
Additional discussion points are in SI sections \ref{SI-sect:vital_rates}-\ref{SI-sect:storage_importance}.

>>>>

%Currently 150 of an allowed 150 words for Ecology Letters
\begin{abstract}
\noindent Species coexistence attracts wide 
interest in ecology. Modern coexistence
theory (MCT) identifies coexistence mechanisms,
one of which, storage effects, hinges on relationships between fluctuations
in environmental and competitive pressures. However, such relationships
are typically measured using covariance, which does not account for 
the possibility that environment and competition may be more related to each other
when they are strong than when weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATAs) are common
between ecological variables, and are important for 
extinction risk, ecosystem stability, and other phenomena. 
We extend MCT, decomposing storage effects to show the influence of ATAs.
Analysis of a simple model
and an empirical example using diatoms
illustrate that ATA influences can be comparable in magnitude to other
mechanisms of coexistence, and that ATAs can make the 
difference between species coexistence and competitive exclusion. 
ATA influences may be an important new mechanism of coexistence. 
\end{abstract}

@@

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has previously been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argue that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting the potential for rescue effects. 
And \cite{Ghosh_2021} show that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples have also been previously explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition and coexistence have not been investigated. Doing so is the goal of this study.

@@

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence species coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition, coexistence, and the maintenance of biodiversity, by means of the 
following specific goals. 
G1) We augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
We examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) We explore, using the simple lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and to determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence and biodiversity maintenance
are likely to be negligible or meaningfully large. 
G3) We provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs relative to other mechanisms of coexistence.
The example is a diatom system in a chemostat. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition and coexistence.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance.

@@

To evaluate the importance of ATAs for a given scenario,
one can quantify
the magnitude of the new ATA effect term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms for the 
scenario.  One typically (but see below) must examine this quantity for each species, $i$,
to obtain complete information about coexistence, since coexistence requires both species populations to 
be able to grow when rare. We also defined scenarios of ``ATA rescue''
and ``ATA exclusion''. \emph{ATA rescue} is considered to have occurred for a given scenario if coexistence
is possible when ATAs are taken into account, i.e., GWR$>0$ for both species, but impossible otherwise, i.e.,
the sum of the non-ATA terms in (\ref{eq:full_decomp}) is negative for at least one species. \emph{ATA exclusion} is 
considered to have occurred if the species do not coexist when ATAs
are taken into account, i.e., GWR$<0$ for at least one species, but would coexist were it not for ATAs, 
i.e., the sum of the non-ATA terms in (\ref{eq:full_decomp})
is positive for both species.

@@

%Setup - two versions, and basics which apply to both
To achieve goal G2 of the Introduction, we applied the framework of Theory to the lottery model, 
comparing the magnitude of the ATA effect on coexistance, $\Delta_i^{[EC]}$, to the magnitudes 
of the other terms of 
the decomposition (\ref{eq:full_decomp}) for various model parameterizations. We used two versions 
of the lottery model. The versions both used the same dynamical equations
(Introduction) and assumed temporally iid fecundities, but the
versions differed in the distributions of $(B_1,B_2)$ used. Both distributions were based on 
transformations of the distributions $(b_1,b_2)$ pictured in Fig. \ref{fig:pedagog1}a-c;
the details of how the left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c), 
and symmetrically (Fig. \ref{fig:pedagog1}b) tail associated alternatives for $(b_1,b_2)$ were generated 
are in SI section \ref{SI-sect:noise}.

@@

%Differences with the original, analytic version of the MCT
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to expand the mathematics of the original MCT to consider ATAs.
Additional discussion points are in SI sections \ref{SI-sect:vital_rates}-\ref{SI-sect:storage_importance}.
